Modelling BCG vaccination in the UK

### 2019-07-12 rmds/03-getTBinR.Rmd
using bookdown::gitbook

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collected by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo[@Abbott:2019] and Github^[GitHub: https://github.com/seabbs/getTBinR/). The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB data set and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used data set before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools. The remainder of this chapter outlines the structure, and key functionality, of `getTBinR 0.6.1` . The use of `getTBinR` in this thesis is explored as well as the use of the package in work external to this thesis and by others. Much of the work done for this chapter was code, documentation, and case study development and so is not fully captured here. The GitHub repository contains a full development history of the package, as well as providing links to the documentation and accompanying case studies.

@@

As `getTBinR` is under active development, the development version can be installed from GitHub with the following (provided the `devtools` package is installed),

```{r gh-installation, eval = FALSE, echo = TRUE}
# install.packages("devtools")
devtools::install_github("seabbs/getTBinR")
```

@@

The data sourced by `getBTinR` is collected by the WHO, via member governments, and used to compile the yearly global TB report [@WHO:2018]. Data collection encompasses TB incidence, TB mortality rates, the age distribution of TB cases, the proportion of drug resistant cases, case detection rates, and treatment rates. For a complete description of the data and data collection process, see [@WHO:2018]. These data are used by the WHO, governmental organisations and researchers to summarise country level TB epidemiology, as well as the wider global and regional picture.

@@

```{r map-tb-incidence-eur, fig.scap = "Map of global TB incidence rates in 2017 as generated by `getTBinR'.", fig.cap = "Map of global TB incidence rates in 2017 as generated by `getTBinR'. Visualising the data with a map allows for spatial trends to be rapidly explored.", echo = TRUE, out.extra = ""}
map_tb_burden(metric = "e_inc_100k", verbose = FALSE)
```

@@

```{r plot-regional-summary, fig.scap = "TB incidence by region and globally as computed and visualised by `getTBinR`.", fig.cap = "TB incidence by region and globally as computed and visualised by `getTBinR`. Confidence intervals have been disabled in order to avoid obscuring the dominant trends.", echo = TRUE, out.extra = ""} 
plot_tb_burden_summary(conf = NULL, metric_label = "e_inc_100k",
                       verbose = FALSE)
```

@@

```{r plot-tb-burden, fig.scap = "TB incidence rates over time, with confidence intervals, in the UK. As produced by `getTBinR`.", fig.cap = "TB incidence rates over time, with confidence intervals, in the UK. As produced by `getTBinR`.", echo = TRUE, out.extra = ""}
plot_tb_burden(metric = "e_inc_100k",
               countries = "United Kingdom",
               verbose = FALSE) 
```

@@

Or accessed online (http://www.seabbs.co.uk/shiny/ExploreGlobalTB). Any metric in the WHO TB data can be explored, with country selection using the built in map, and animation possible by year. The shiny app can also be used to generate the country level reports discussed in the next section. Figure \@ref(fig:dash-snapshot) shows a screenshot of the dashboard, with South Africa selected as the country of interest.

@@

As of the 1st of August 2019, `getTBinR` has been released on CRAN for over a year. It has been downloaded over 10,000 times, has a growing user base, no outstanding bug-related issues, and has received multiple updates greatly expanding the functionality available. The standalone dashboard hosted online^[Dashboard: http://www.seabbs.co.uk/shiny/ExploreGlobalTB/] has had over 3000 unique users. It has been used as a teaching aid, as an example of open science, to facilitate exploratory data analysis and to provide context for other research. In this thesis it has been used extensively in Chapter \@ref(background) to provide context and was also used as a hypothesis generating tool in all other chapters. Outside of the work presented in this thesis I have used `getTBinR` extensively as a data analysis tool^[Blog: https://www.samabbott.co.uk/tags/who/], to widen the appreciation of TB as a global health problem^[Twitter:https://twitter.com/search?q=getTBinR&src=typd]^[Reddit: https://www.reddit.com/user/seabbs/posts/], and to provide contextual information for other research^[Presentation: https://www.samabbott.co.uk/what-do-we-really-know-about-bcg/presentation.html#1].

Whilst `getTBinR` is feature complete, and stable, development work continues. Future projects include: using the `shinymeta` R package to provide downloadable R code to users of the interactive application; iterating on the current automated report to improve formatting and to increase the amount of information displayed. As additional WHO TB data are released they will be added to `getTBinR`.

@@

 * The current impact and future direction of `getTBinR` has also be detailed.

>>>>

`getTBinR` is an R package[@RCoreTeam2019] that I developed to facilitate working with the data[@WHO:2018] collected by the World Health Organization (WHO) on the country level epidemiology of Tuberculosis (TB). All data is freely available from the WHO^[WHO TB data: https://www.who.int/tb/country/data/download/en/] and the package code is archived on Zenodo[@Abbott:2019] and Github^[GitHub: https://github.com/seabbs/getTBinR/]. The aim of `getTBinR` is to allow researchers, and other interested individuals, to quickly and easily gain access to a detailed TB data set and to start using it to derive key insights. It provides a consistent set of tools that can be used to rapidly evaluate hypotheses on a widely used data set before they are explored further using more complex methods or more detailed data. Prior to the development of `getTBinR` access to the WHO data was ad-hoc and there were no standardised visualisation or summary tools.

The remainder of this chapter outlines the structure, and key functionality, of `getTBinR 0.6.1` . The use of `getTBinR` in this thesis is explored as well as the use of the package in work external to this thesis and by others. Much of the work done for this chapter was code, documentation, and case study development and so is not fully captured here. The GitHub repository contains a full development history of the package, as well as providing links to the documentation and accompanying case studies.

@@

As `getTBinR` is under active development, the development version can be installed from GitHub with the following,

```{r gh-installation, eval = FALSE, echo = TRUE}
# install.packages("remotes")
remotes::install_github("seabbs/getTBinR")
```

@@

The data sourced by `getBTinR` is collected by the WHO, via member governments, and used to compile the yearly global TB report.[@WHO:2018] Data collection encompasses TB incidence, TB mortality rates, the age distribution of TB cases, the proportion of drug resistant cases, case detection rates, and treatment rates. For a complete description of the data and data collection process, see [@WHO:2018]. These data are used by the WHO, governmental organisations and researchers to summarise country level TB epidemiology, as well as the wider global and regional picture.

@@

```{r map-tb-incidence-eur, fig.scap = "Map of global TB incidence rates in 2017 as generated by getTBinR.", fig.cap = "Map of global TB incidence rates in 2017 as generated by getTBinR. Visualising the data with a map allows for spatial trends to be rapidly explored.", echo = TRUE, out.extra = ""}
map_tb_burden(metric = "e_inc_100k", verbose = FALSE)
```

@@

```{r plot-regional-summary, fig.scap = "TB incidence by region and globally as computed and visualised by getTBinR.", fig.cap = "TB incidence by region and globally as computed and visualised by getTBinR. Confidence intervals have been disabled in order to avoid obscuring the dominant trends.", echo = TRUE, out.extra = ""} 
plot_tb_burden_summary(conf = NULL, metric_label = "e_inc_100k",
                       verbose = FALSE)
```

@@

```{r plot-tb-burden, fig.scap = "TB incidence rates over time, with confidence intervals, in the UK. As produced by getTBinR.", fig.cap = "TB incidence rates over time, with confidence intervals, in the UK. As produced by getTBinR.", echo = TRUE, out.extra = ""}
plot_tb_burden(metric = "e_inc_100k",
               countries = "United Kingdom",
               verbose = FALSE) 
```

@@

Or accessed online^[Dashboard: http://www.seabbs.co.uk/shiny/ExploreGlobalTB]. Any metric in the WHO TB data can be explored, with country selection using the built in map, and animation possible by year. The shiny app can also be used to generate the country level reports discussed in the next section. Figure \@ref(fig:dash-snapshot) shows a screenshot of the dashboard, with South Africa selected as the country of interest.

@@

As of the 1st of August 2019, `getTBinR` has been released on CRAN for over a year. It has been downloaded over 10,000 times, has a growing user base, no outstanding bug-related issues, and has received multiple updates greatly expanding the functionality available. The standalone dashboard hosted online^[Dashboard: http://www.seabbs.co.uk/shiny/ExploreGlobalTB/] has had over 3000 unique users. It has been used as a teaching aid, as an example of open science, to facilitate exploratory data analysis and to provide context for other research. In this thesis it has been used extensively in Chapter \@ref(background) to provide context and was also used as a hypothesis generating tool in all other chapters. Outside of the work presented in this thesis I have used `getTBinR` extensively as a data analysis tool^[Blog: https://www.samabbott.co.uk/tags/who/], to widen the appreciation of TB as a global health problem^[Twitter: https://twitter.com/search?q=getTBinR&src=typd; Reddit: https://www.reddit.com/user/seabbs/posts/], and to provide contextual information for other research^[Presentation: https://www.samabbott.co.uk/what-do-we-really-know-about-bcg/presentation.html#1].

Whilst `getTBinR` is feature complete, and stable, development work continues. Future projects include: using the `shinymeta` R package to provide downloadable R code to users of the interactive application; iterating on the current automated report to improve formatting and to increase the amount of information displayed, and expanding the range of visualisation functions available. As additional WHO TB data are released they will be added to `getTBinR`.

@@

 * The current impact and future direction of `getTBinR` has also been detailed.

### 2019-07-12 rmds/04-tb-epi-england.Rmd
using bookdown::gitbook

In this chapter I explore the epidemiology of TB in England using routine data-sets, with a particular focus on the impact of missing data, the mechanisms underlying that missing date, seasonal trends, the role of age, UK birth status and BCG status. I have also estimated incidence rates, stratified by UK birth status and age, which I then used to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data will then be used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(benefical-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fit)).

@@

Findings from this chapter are used throughout the later chapters of this thesis. In particular, Chapter  \@ref(benefical-bcg-out) uses statistical modelling to exploring the impact of BCG vaccination on TB outcomes in greater detail, Chapter \@ref(direct-eff) explores the impact of the change in BCG vaccination policy on TB incidence rates using the incidence rate estimates from this chapter, Chapter \@ref(model-development) uses the understanding of the ETS gained from this chapter to parameterise a dynamic TB transmission model, and Chapter \@ref(model-fit) uses the insights gained into the date variables in the ETS to fit a dynamic TB transmission model.

@@

- Findings from this chapter are used throughout the later chapters of this thesis; to inform analysis questions (Chapter  \@ref(benefical-bcg-out) and \@ref(direct-eff)), identify variables for which missing data must be imputed (Chapter  \@ref(benefical-bcg-out) and \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)), and to fit a dynamic TB transmission model (Chapter \@ref(model-fit)).

>>>>

In this chapter I explore the epidemiology of TB in England using routine data-sets, with a particular focus on the impact of missing data, the mechanisms underlying that missing date, seasonal trends, the role of age, UK birth status and BCG status. I have also estimated incidence rates, stratified by UK birth status and age, which I then used to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data will then be used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(benefical-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fitting)).

@@

Findings from this chapter are used throughout the later chapters of this thesis. In particular, Chapter  \@ref(benefical-bcg-out) uses statistical modelling to exploring the impact of BCG vaccination on TB outcomes in greater detail, Chapter \@ref(direct-eff) explores the impact of the change in BCG vaccination policy on TB incidence rates using the incidence rate estimates from this chapter, Chapter \@ref(model-development) uses the understanding of the ETS gained from this chapter to parameterise a dynamic TB transmission model, and Chapter \@ref(model-fitting) uses the insights gained into the date variables in the ETS to fit a dynamic TB transmission model.

@@

- Findings from this chapter are used throughout the later chapters of this thesis; to inform analysis questions (Chapter  \@ref(benefical-bcg-out) and \@ref(direct-eff)), identify variables for which missing data must be imputed (Chapter  \@ref(benefical-bcg-out) and \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)), and to fit a dynamic TB transmission model (Chapter \@ref(model-fitting)).

### 2019-07-12 rmds/index.Rmd
using bookdown::gitbook

---
author: 'Samuel Abbott'
date: 'August 2019'
division: 'Bristol Medical School\: Population Health Sciences'
advisor: 'Ellen Brooks Pollock'
altadvisor: 'Hannah Christensen'
department: 'Faculty of Health Sciences'
title: 'Modelling BCG vaccination in the UK'
subtitle: 'What is the impact of changing policy?'
wordcount: "41,218"
knit: bookdown::render_book
site: bookdown::bookdown_site
always_allow_html: yes
output: 
  #thesisdown::thesis_pdf: default
  thesisdown::thesis_gitbook: default
  #thesisdown::thesis_word: default
  #thesisdown::thesis_epub: default
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = ' ')`
acknowledgements: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-acknowledgements.Rmd"), collapse = ' ')`
declaration: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-declaration.Rmd"), collapse = ' ')`
contribution: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-contribution.Rmd"), collapse = ' ')`
acronyms: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-acronyms.Rmd"), collapse = ' ')`
bibliography: bib/thesis.bib
#Download your specific bibliography database file and refer to it in the line above.
csl: csl/bmj.csl
#Download your specific csl file and refer to it in the line above.
lot: true
lof: true
space_between_paragraphs: true
#Delete the # at the beginning of the previous line if you'd like
#to have a blank new line between each paragraph
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}

>>>>

---
author: 'Samuel Abbott'
date: 'August 2019'
division: 'Bristol Medical School\: Population Health Sciences'
advisor: 'Ellen Brooks Pollock'
altadvisor: 'Hannah Christensen'
department: 'Faculty of Health Sciences'
title: 'Modelling BCG vaccination in the UK'
subtitle: 'What is the impact of changing policy?'
wordcount: "41,218"
knit: bookdown::render_book
site: bookdown::bookdown_site
always_allow_html: yes
output: 
  #thesisdown::thesis_pdf: default
  bookdown::render_book: default
  #thesisdown::thesis_word: default
  #thesisdown::thesis_epub: default
abstract: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-abstract.Rmd"), collapse = ' ')`
acknowledgements: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-acknowledgements.Rmd"), collapse = ' ')`
declaration: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-declaration.Rmd"), collapse = ' ')`
contribution: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-contribution.Rmd"), collapse = ' ')`
acronyms: |
  `r if(knitr:::is_latex_output()) paste(readLines("00-acronyms.Rmd"), collapse = ' ')`
bibliography: [bib/thesis.bib]
#Download your specific bibliography database file and refer to it in the line above.
link-citations: yes
colorlinks: yes
csl: [csl/bmj.csl]
#Download your specific csl file and refer to it in the line above.
lot: true
lof: true
space_between_paragraphs: true
#Delete the # at the beginning of the previous line if you'd like
#to have a blank new line between each paragraph
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage[table]{xcolor}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
